---
title: "CD59 expression and overall survival in the BeatAML2 cohort"
author: "Abdula Maher"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

<!--
========================================================================
  CD59 / BeatAML2 survival analysis

  Cell-surface (cluster-of-differentiation) gene expression was screened
  against overall survival in the BeatAML2 cohort (waves 1-4, initial
  diagnosis specimens). CD59 emerged as a candidate prognostic marker;
  this notebook reproduces the supporting analyses.

  DATA ACCESS
  -----------
  BeatAML2 expression, clinical, and mutation tables are distributed
  through Vizome (https://biodev.github.io/BeatAML2/) and dbGaP
  (phs001657). Genomic tables are controlled-access. Place the four
  input files in the directory pointed to by `data_dir` below. No
  patient-level data are contained in this script.

  REPRODUCIBILITY NOTES
  ---------------------
  * File paths are parameterised (`data_dir`); no absolute/personal
    paths are hard-coded.
  * CD59 expression cut-offs are computed from the data (25th/75th
    percentile), not hard-coded.
  * Group memberships are selected by column NAME rather than by
    positional index, so the pipeline is robust to column re-ordering.
  * A handful of schema-dependent steps are flagged with `## SCHEMA:`
    and should be verified against your copy of the clinical table on
    first run (BeatAML column positions can shift between releases).
========================================================================
-->

# 1. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install (if needed) and load packages. CRAN and Bioconductor packages are
handled separately.

```{r packages}
cran_pkgs <- c(
  "survival", "survminer", "ggrepel", "ggpubr", "dplyr", "tidyr",
  "tibble", "ggprism", "aorsf", "here"
)
bioc_pkgs <- c(
  "GSVA", "GSEABase", "limma", "fgsea", "ComplexHeatmap", "pheatmap"
)

new_cran <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(new_cran)) install.packages(new_cran, dependencies = TRUE)

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
new_bioc <- bioc_pkgs[!bioc_pkgs %in% rownames(installed.packages())]
if (length(new_bioc)) BiocManager::install(new_bioc, update = FALSE, ask = FALSE)

suppressPackageStartupMessages({
  library(survival)
  library(survminer)
  library(ggrepel)
  library(ggpubr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggprism)
})
```

## 1.1 Analysis parameters and helper functions

Set the location of the input files and define a few small helpers that are
reused throughout the notebook.

```{r params-helpers}
# ---- Location of the four BeatAML2 input files ----------------------------
# Override by setting the BEATAML2_DIR environment variable, or edit here.
data_dir <- Sys.getenv("BEATAML2_DIR", unset = ".")

# ---- Genes / signatures used repeatedly -----------------------------------
top_mutated_genes <- c(
  "DNMT3A", "NRAS", "TET2", "NPM1", "IDH2", "RUNX1", "SRSF2", "TP53",
  "FLT3", "IDH1", "ASXL1", "STAG2", "KRAS", "U2AF1", "WT1"
)

# ---- Helper: assign expression tertile labels from fixed cut-offs ----------
# Values >= `high` -> "Top 25", <= `low` -> "Bottom 25", else "Median 50".
assign_quartile <- function(x, low, high,
                            labels = c("Bottom 25", "Median 50", "Top 25")) {
  x <- as.numeric(x)
  grp <- ifelse(x >= high, labels[3],
         ifelse(x <= low,  labels[1], labels[2]))
  factor(grp, levels = labels)
}

# ---- Helper: standard CD59 expression box-plot vs a categorical variable ---
cd59_group_boxplot <- function(df, xvar, xlab, title,
                               fill_values = NULL,
                               method = "t.test",
                               ref_group = NULL) {
  p <- ggplot(df, aes(x = .data[[xvar]], y = CD59, fill = .data[[xvar]])) +
    geom_boxplot() +
    theme_prism() +
    stat_compare_means(
      aes(label = after_stat(p.signif)),
      method    = method,
      ref.group = ref_group,
      size      = 6
    ) +
    labs(x = xlab, y = "CD59 expression (z-score)", title = title) +
    theme(legend.position = "none")
  if (!is.null(fill_values)) p <- p + scale_fill_manual(values = fill_values)
  p
}

# ---- Helper: Wilcoxon comparison of many signatures between two groups -----
# Returns a data.frame with delta (group2 - group1), p-value and BH-FDR.
# Signatures with fewer than `min_n` non-missing values in either group return
# an NA p-value instead of aborting (e.g. a signature whose genes are absent
# from the expression matrix, which would otherwise crash wilcox.test).
compare_signatures <- function(df, group_col, sig_cols, group1, group2,
                               min_n = 3) {
  res <- lapply(sig_cols, function(s) {
    x1 <- as.numeric(df[[s]][df[[group_col]] == group1])
    x2 <- as.numeric(df[[s]][df[[group_col]] == group2])
    x1 <- x1[is.finite(x1)]      # drop NA and NaN
    x2 <- x2[is.finite(x2)]
    p <- if (length(x1) >= min_n && length(x2) >= min_n) {
      tryCatch(wilcox.test(x2, x1)$p.value, error = function(e) NA_real_)
    } else {
      NA_real_
    }
    data.frame(
      signature   = s,
      n_group1    = length(x1),
      n_group2    = length(x2),
      mean_group1 = if (length(x1)) mean(x1) else NA_real_,
      mean_group2 = if (length(x2)) mean(x2) else NA_real_,
      delta_mean  = (if (length(x2)) mean(x2) else NA_real_) -
                    (if (length(x1)) mean(x1) else NA_real_),
      p_value     = p,
      stringsAsFactors = FALSE
    )
  })
  res <- do.call(rbind, res)
  res$FDR <- p.adjust(res$p_value, method = "BH")   # NA p-values stay NA
  res[order(res$FDR), ]
}
```

# 2. Load data

```{r load-data}
harm_data <- read.delim(
  file.path(data_dir, "beataml_waves1to4_norm_exp_dbgap.txt"),
  stringsAsFactors = FALSE, header = TRUE
)
clinical_data <- read.csv(
  file.path(data_dir, "beataml_wv1to4_clinical.csv"),
  header = TRUE, check.names = FALSE
)
seq_data <- read.delim(
  file.path(data_dir, "beataml_wes_wv1to4_mutations_dbgap.txt"),
  stringsAsFactors = FALSE, header = TRUE
)
```

# 3. Expression normalisation (z-scores)

Quantile-normalised expression is converted to per-gene z-scores.

```{r zscores}
harm_data <- unique(as.data.frame(harm_data))

# The first four columns are gene annotation (stable id, display label,
# description, biotype); the remainder are per-sample expression values.
rownames(harm_data) <- make.names(harm_data$display_label, unique = TRUE)
annotation_cols <- 1:4                     ## SCHEMA: annotation columns
harm_data <- harm_data[, -annotation_cols]

# Per-gene z-score: scale() works on columns, so transpose in and out.
zscore_counts <- t(scale(t(harm_data)))
rownames(zscore_counts) <- rownames(harm_data)
```

```{r zscore-qc}
hist(zscore_counts,
     breaks = 100,
     main = "Distribution of expression z-scores",
     xlab = "z-score")
```

# 4. Build the analysis table (`survplotdata`)

Clinical variables are joined to the z-scored expression matrix, keeping only
initial-diagnosis samples that have RNA-seq data and complete values for the
clinical covariates used downstream.

```{r build-metadata}
clinical_filtered <- clinical_data %>%
  filter(
    dbgap_rnaseq_sample != "",
    diseaseStageAtSpecimenCollection == "Initial Diagnosis"
  )

# "TP53" and "ASXL1" here are the clinical mutation-consensus calls. They share
# their names with the TP53/ASXL1 expression columns, so rename them to
# "<gene>_mut" BEFORE the merge to avoid a name collision (cbind would otherwise
# create two identically-named columns and the mutation calls become unreachable).
clinical_vars <- c(
  "ELN2017", "vitalStatus", "overallSurvival", "TP53", "ASXL1",
  "isRelapse", "isDenovo", "priorMDS", "responseToInductionTx",
  "responseDurationToInductionTx", "ageAtDiagnosis", "wbcCount"
)

metadata <- clinical_filtered[, clinical_vars, drop = FALSE]
rownames(metadata) <- clinical_filtered$dbgap_rnaseq_sample

names(metadata)[names(metadata) == "TP53"]  <- "TP53_mut"
names(metadata)[names(metadata) == "ASXL1"] <- "ASXL1_mut"

# Drop samples with a missing value in any selected clinical covariate.
metadata <- metadata[complete.cases(metadata), , drop = FALSE]

# Align expression and clinical tables on a shared, explicitly-ordered set of
# sample IDs (avoids relying on incidental row ordering).
common_ids <- intersect(rownames(metadata), colnames(zscore_counts))
metadata   <- metadata[common_ids, , drop = FALSE]
expr_t     <- as.data.frame(t(zscore_counts[, common_ids, drop = FALSE]))

coxdata <- cbind(expr_t, metadata)

# Guard against any remaining accidental duplicate column names.
stopifnot(!anyDuplicated(names(coxdata)))
```

```{r type-coercion}
coxdata$overallSurvival <- as.numeric(coxdata$overallSurvival)
coxdata$ageAtDiagnosis  <- as.numeric(coxdata$ageAtDiagnosis)
coxdata$wbcCount        <- as.numeric(coxdata$wbcCount)
coxdata$ELN2017         <- factor(coxdata$ELN2017)
coxdata$isRelapse       <- factor(coxdata$isRelapse)

# Encode vital status as a 0/1 event indicator and drop "Unknown".
coxdata <- coxdata[coxdata$vitalStatus != "Unknown", , drop = FALSE]
coxdata$vitalStatus <- dplyr::case_when(
  coxdata$vitalStatus == "Dead"  ~ 1L,
  coxdata$vitalStatus == "Alive" ~ 0L,
  TRUE ~ NA_integer_
)
coxdata <- coxdata[!is.na(coxdata$vitalStatus), , drop = FALSE]

# Normalise mutation-consensus columns to Wildtype / Mutation.
to_status <- function(x) {
  x[is.na(x) | x %in% c("", "negative")] <- "Wildtype"
  x[x != "Wildtype"] <- "Mutation"
  factor(x, levels = c("Wildtype", "Mutation"))
}
coxdata$TP53_mut  <- to_status(coxdata$TP53_mut)
coxdata$ASXL1_mut <- to_status(coxdata$ASXL1_mut)
```

Assemble `survplotdata`: all CD-gene expression columns plus the clinical
covariates, selected by name.

```{r survplotdata}
cd_genes_file <- read.csv(file.path(data_dir, "CD_genes.csv"))
cd_gene_symbols <- cd_genes_file$Approved.symbol

# Keep only CD genes that are present as expression columns.
cd_gene_cols <- intersect(cd_gene_symbols, colnames(coxdata))

clinical_keep <- c(
  "ELN2017", "vitalStatus", "overallSurvival", "TP53_mut", "ASXL1_mut",
  "isRelapse", "isDenovo", "priorMDS", "responseToInductionTx",
  "responseDurationToInductionTx", "ageAtDiagnosis", "wbcCount"
)
clinical_keep <- intersect(clinical_keep, colnames(coxdata))

survplotdata <- coxdata[, c(cd_gene_cols, clinical_keep), drop = FALSE]

# Tidy the one known label typo in the response variable.
survplotdata$responseToInductionTx[
  survplotdata$responseToInductionTx == "Complete Reponse i"
] <- "Complete Response"
```

# 5. Define CD59 expression groups (once)

CD59 tertile membership (bottom 25% / middle 50% / top 25%) is computed from
the cohort's own 25th and 75th percentiles and reused everywhere.

```{r cd59-groups}
survplotdata$CD59 <- as.numeric(survplotdata$CD59)

cd59_cuts <- quantile(survplotdata$CD59, probs = c(0.25, 0.75), na.rm = TRUE)
lowExpr  <- cd59_cuts[["25%"]]
highExpr <- cd59_cuts[["75%"]]

survplotdata$CD59quartile <- assign_quartile(survplotdata$CD59, lowExpr, highExpr)

table(survplotdata$CD59quartile)

top25_ids    <- rownames(survplotdata)[survplotdata$CD59quartile == "Top 25"]
bottom25_ids <- rownames(survplotdata)[survplotdata$CD59quartile == "Bottom 25"]
```

# 6. Per-sample mutation status (for CD59-high vs -low comparison)

Mutation status for the recurrently mutated genes is derived from the WES
mutation table by sample and gene symbol (name-based, so robust to column
order).

```{r mutation-status}
clinical_dna <- clinical_filtered %>% filter(dbgap_dnaseq_sample != "")

seq_dna <- seq_data %>%
  filter(dbgap_sample_id %in% clinical_dna$dbgap_dnaseq_sample,
         symbol %in% top_mutated_genes) %>%
  distinct(dbgap_sample_id, symbol)

# Initialise every gene to Wildtype, then flip to Mutation where a call exists.
for (g in top_mutated_genes) {
  mutated_samples <- seq_dna$dbgap_sample_id[seq_dna$symbol == g]
  clinical_dna[[paste0(g, "_mut")]] <- ifelse(
    clinical_dna$dbgap_dnaseq_sample %in% mutated_samples,
    "Mutation", "Wildtype"
  )
}

# Assemble a per-RNA-seq-sample mutation table aligned to survplotdata,
# carrying CD59 expression so groups can be defined the same way as above.
mut_cols <- paste0(top_mutated_genes, "_mut")
mutation_data <- clinical_dna %>%
  select(dbgap_rnaseq_sample, all_of(mut_cols)) %>%
  filter(dbgap_rnaseq_sample %in% rownames(survplotdata))
rownames(mutation_data) <- mutation_data$dbgap_rnaseq_sample

mutation_data$CD59 <- survplotdata[rownames(mutation_data), "CD59"]
mutation_data$CD59quartile <-
  assign_quartile(mutation_data$CD59, lowExpr, highExpr)
```

# 7. CD59 expression across clinical / mutational features

The four comparisons below share one plotting helper (`cd59_group_boxplot`).

```{r cd59-vs-tp53}
cd59_group_boxplot(
  survplotdata, "TP53_mut",
  xlab  = "TP53 mutation status",
  title = "CD59 expression vs TP53 mutation status",
  fill_values = c("blue", "red")
)
```

```{r cd59-vs-mds}
cd59_group_boxplot(
  survplotdata, "priorMDS",
  xlab  = "Prior MDS",
  title = "CD59 expression vs prior MDS",
  fill_values = c("skyblue", "red3")
)
```

```{r cd59-vs-response}
resp_df <- subset(
  survplotdata,
  responseToInductionTx %in% c("Complete Response", "Refractory")   # %in%, not ==
)
cd59_group_boxplot(
  resp_df, "responseToInductionTx",
  xlab  = "Response to induction treatment",
  title = "CD59 expression vs response to induction",
  fill_values = c("blue", "red")
)
```

```{r cd59-vs-eln}
survplotdata$ELN2017 <- factor(
  survplotdata$ELN2017,
  levels = c("Favorable", "Intermediate", "FavorableOrIntermediate",
             "Adverse", "IntermediateOrAdverse", "MissingKaryo",
             "NonAML", "NonInitial")
)

eln_df <- subset(
  survplotdata,
  ELN2017 %in% c("Favorable", "Intermediate", "Adverse")            # %in%, not ==
)
eln_df$ELN2017 <- droplevels(eln_df$ELN2017)

cd59_group_boxplot(
  eln_df, "ELN2017",
  xlab  = "ELN 2017 classification",
  title = "CD59 expression vs ELN 2017 classification",
  fill_values = c("gray", "blue", "red"),
  ref_group = "Favorable"
) +
  scale_y_continuous(limits = c(-3, 2.4))
```

# 8. Multivariable Cox model and forest plot

CD59 tertile adjusted for ELN 2017, age, WBC count, and de novo status.
`Surv()` is written inline so `ggforest()` can rebuild the model frame.

```{r cox-multivariable}
# Attach the CD59 tertile to a frame carrying the model covariates.
cox_df <- survplotdata[, c(
  "overallSurvival", "vitalStatus", "CD59quartile",
  "ELN2017", "ageAtDiagnosis", "wbcCount", "isDenovo"
)]
cox_df$isDenovo <- factor(cox_df$isDenovo)

fit_cox_mv <- coxph(
  Surv(overallSurvival, vitalStatus) ~
    CD59quartile + ELN2017 + ageAtDiagnosis + wbcCount + isDenovo,
  data = cox_df
)

summary(fit_cox_mv)
cat("N =", fit_cox_mv$n, " events =", fit_cox_mv$nevent, "\n")

ggforest(fit_cox_mv, data = cox_df)
```

```{r}
library(ggplot2)
library(patchwork)

forest_plot <- function(df,
                        effect_label = "Hazard ratio",
                        ci_header    = "HR (95% CI)",
                        stat_header  = "N (events)",
                        log_scale    = TRUE,
                        ref_line     = 1,
                        y_face       = "italic") {

  df <- df[nrow(df):1, ]            # first row at the top
  df$y <- seq_len(nrow(df)); n <- nrow(df)

  stars <- function(p) ifelse(is.na(p), "",
                       ifelse(p < 0.001, "***",
                       ifelse(p < 0.01,  "**",
                       ifelse(p < 0.05,  "*", ""))))
  df$star  <- stars(df$p.value)
  is_ref   <- is.na(df$conf.low)
  df$hr_ci <- ifelse(is_ref, "1.00 (reference)",
                     sprintf("%.2f (%.2f\u2013%.2f)", df$estimate, df$conf.low, df$conf.high))
  df$p_fmt <- ifelse(is.na(df$p.value), "",
                     ifelse(df$p.value < 0.001, "<0.001",
                            formatC(df$p.value, format = "f", digits = 3)))
  header_y <- n + 0.9

  p_forest <- ggplot(df, aes(estimate, y)) +
    geom_vline(xintercept = ref_line, colour = "grey55", linewidth = 0.4) +
    geom_errorbarh(data = function(d) d[!is.na(d$conf.low), ],
                   aes(xmin = conf.low, xmax = conf.high),
                   height = 0.16, linewidth = 0.6) +
    geom_point(shape = 15, size = 3.4) +
    geom_text(aes(label = star), vjust = -0.9, size = 5) +
    scale_x_continuous(trans = if (log_scale) "log10" else "identity") +
    scale_y_continuous(labels = df$label, breaks = df$y,
                       limits = c(0.5, header_y + 0.3), expand = c(0, 0)) +
    labs(x = effect_label, y = NULL) +
    theme_minimal(base_size = 13) +
    theme(panel.grid    = element_blank(),
          axis.line.x   = element_line(colour = "black", linewidth = 0.4),
          axis.ticks.x  = element_line(colour = "black", linewidth = 0.4),
          axis.text.y   = element_text(face = y_face, hjust = 0, colour = "black"),
          axis.ticks.y  = element_blank())

  xc <- c(1, 2.15, 3.1)
  p_table <- ggplot(df, aes(y = y)) +
    geom_text(aes(x = xc[1], label = hr_ci), size = 4.2) +
    geom_text(aes(x = xc[2], label = p_fmt), size = 4.2) +
    geom_text(aes(x = xc[3], label = stat),  size = 4.2) +
    annotate("text", x = xc, y = header_y,
             label = c(ci_header, "P value", stat_header), size = 4.4) +
    annotate("segment", x = xc - 0.42, xend = xc + 0.42,
             y = header_y - 0.45, yend = header_y - 0.45, linewidth = 0.4) +
    scale_x_continuous(limits = c(0.4, 3.7), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0.5, header_y + 0.3), expand = c(0, 0)) +
    theme_void()

  p_forest + p_table + plot_layout(widths = c(1, 1.05))
}
```

```{r}
library(broom)

# (tidy_with_ref() definition stays here, unchanged)

# Restrict ELN to the three clean categories and refit.
cox_df3 <- subset(cox_df, ELN2017 %in% c("Favorable", "Intermediate", "Adverse"))
cox_df3$ELN2017      <- factor(cox_df3$ELN2017, levels = c("Favorable", "Intermediate", "Adverse"))
cox_df3$CD59quartile <- factor(cox_df3$CD59quartile, levels = c("Bottom 25", "Median 50", "Top 25"))
cox_df3$isDenovo     <- relevel(factor(cox_df3$isDenovo), ref = "TRUE")   # de novo AML = reference

fit_cox_mv3 <- coxph(
  Surv(overallSurvival, vitalStatus) ~
    CD59quartile + ELN2017 + ageAtDiagnosis + isDenovo,   # WBC removed
  data = cox_df3)

pretty  <- c(CD59quartile = "CD59", ELN2017 = "ELN", ageAtDiagnosis = "Age",
             isDenovo = "De novo")
plot_df <- tidy_with_ref(fit_cox_mv3, cox_df3, pretty)

forest_plot(plot_df, effect_label = "Hazard ratio", stat_header = "n", y_face = "plain")
ggsave("cd59_forest.pdf", width = 10, height = 4.6)
```


# 9. Mutation enrichment: CD59-high vs CD59-low (Fisher)

Per-gene Fisher exact tests comparing mutation frequency between the CD59 top
and bottom tertiles.

```{r fisher-enrichment}
mut_matrix <- function(quartile) {
  m <- mutation_data[mutation_data$CD59quartile == quartile, mut_cols, drop = FALSE]
  # Rows = genes, entries logical (TRUE = mutated).
  t(m == "Mutation")
}

high_mut <- mut_matrix("Top 25")
low_mut  <- mut_matrix("Bottom 25")
rownames(high_mut) <- sub("_mut$", "", rownames(high_mut))
rownames(low_mut)  <- sub("_mut$", "", rownames(low_mut))

genes  <- intersect(rownames(high_mut), rownames(low_mut))
n_high <- ncol(high_mut)
n_low  <- ncol(low_mut)

fisher_results <- lapply(genes, function(g) {
  a <- sum(high_mut[g, ], na.rm = TRUE)   # high, mutated
  b <- n_high - a                         # high, wildtype
  c <- sum(low_mut[g, ],  na.rm = TRUE)   # low,  mutated
  d <- n_low - c                          # low,  wildtype
  ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2, byrow = TRUE))
  data.frame(
    gene = g, high_mut = a, high_wt = b, low_mut = c, low_wt = d,
    high_freq = a / n_high, low_freq = c / n_low,
    delta_freq = a / n_high - c / n_low,
    odds_ratio = unname(ft$estimate), p_value = ft$p.value,
    stringsAsFactors = FALSE
  )
})
fisher_results <- do.call(rbind, fisher_results)
fisher_results$FDR <- p.adjust(fisher_results$p_value, method = "BH")

# Drop very rare events (unstable tests) and order by significance.
fisher_results <- fisher_results %>%
  filter((high_mut + low_mut) >= 3) %>%
  arrange(FDR, desc(delta_freq))

head(fisher_results, 20)

write.csv(fisher_results, "CD59_High_vs_Low_Fisher.csv", row.names = FALSE)
```

# 10. Pan-CD-gene hazard-ratio screen

For every CD gene, split patients into expression tertiles and fit a
univariable Cox model; record the top-vs-bottom hazard ratio and p-value.
Coefficients are extracted by NAME (robust to genes that lack one tertile).

```{r cd-gene-quartiles}
# Assign tertiles for each CD gene. NAs are left as NA (patient excluded from
# that gene's model) rather than imputed to the mean.
quartile_data <- survplotdata
for (g in cd_gene_cols) {
  cuts <- quantile(as.numeric(quartile_data[[g]]),
                   probs = c(0.25, 0.75), na.rm = TRUE)
  quartile_data[[paste0(g, "quartile")]] <-
    assign_quartile(quartile_data[[g]], cuts[[1]], cuts[[2]])
}

quartile_cols <- paste0(cd_gene_cols, "quartile")
```

```{r cd-gene-hr-screen}
hz_list <- lapply(cd_gene_cols, function(g) {
  qcol <- paste0(g, "quartile")
  fit  <- coxph(
    as.formula(sprintf("Surv(overallSurvival, vitalStatus) ~ `%s`", qcol)),
    data = quartile_data
  )
  sc  <- summary(fit)$coefficients
  row <- grep("Top 25$", rownames(sc), value = TRUE)   # top-vs-bottom term
  data.frame(
    gene   = g,
    Hazard = if (length(row) == 1) sc[row, "exp(coef)"] else NA_real_,
    PVal   = if (length(row) == 1) sc[row, "Pr(>|z|)"] else NA_real_,
    stringsAsFactors = FALSE
  )
})
hz_df <- do.call(rbind, hz_list)
hz_df$FDR <- p.adjust(hz_df$PVal, method = "BH")

write.csv(hz_df, "HR_CD_Gene_Quartile.csv", row.names = FALSE)
head(hz_df[order(hz_df$FDR), ], 15)
```

```{r cd-gene-hr-volcano}
label_df <- subset(hz_df, Hazard > 1.7 & !is.na(Hazard))

ggplot(hz_df, aes(x = Hazard, y = -log10(PVal))) +
  geom_point(size = 3) +
  geom_point(data = label_df, colour = "blue", size = 3) +
  geom_point(data = subset(hz_df, gene == "CD59"),
             colour = "red", size = 3) +
  geom_text_repel(
    data = label_df, aes(label = gene),
    size = 2.5, max.overlaps = 20, nudge_x = 0.1, nudge_y = 0.1
  ) +
  scale_x_continuous(limits = c(0, 4)) +
  scale_y_continuous(limits = c(0, 8)) +
  theme_prism() +
  labs(
    title = "Hazard ratios of cluster-of-differentiation genes",
    x = "Hazard ratio (top vs bottom tertile)",
    y = "-log10(p-value)"
  )
```

# 11. Kaplan-Meier: overall survival by CD59 tertile

```{r kaplan-meier}
fit_km <- survfit(
  Surv(overallSurvival, vitalStatus) ~ CD59quartile,
  data = survplotdata
)

ggsurvplot(
  fit_km, data = survplotdata,
  pval = TRUE, conf.int = FALSE, risk.table = FALSE,
  ggtheme = theme_prism(),
  legend.labs  = c("Bottom 25", "Median 50", "Top 25"),
  legend.title = "CD59 expression",
  palette = c("blue", "#464747", "red"),
  title = "Overall survival vs CD59 expression tertile",
  xlab  = "Overall survival (days)",
  ylab  = "Survival probability"
)
```

# 12. Oblique random survival forest (aorsf)

An oblique RSF using CD59 tertile; expected risk over time is read out via
out-of-bag partial dependence. (Additional covariates can be added to the
formula if a multivariable forest is desired.)

```{r aorsf}
library(aorsf)

fit_orsf <- orsf(
  Surv(overallSurvival, vitalStatus) ~ CD59quartile,
  data = survplotdata
)

orsf_pd_oob(
  fit_orsf,
  pred_spec    = list(CD59quartile = c("Bottom 25", "Median 50", "Top 25")),
  pred_horizon = seq_len(600)
) %>%
  ggplot(aes(x = pred_horizon, y = mean, colour = CD59quartile)) +
  geom_line(linewidth = 1) +
  labs(x = "Time since baseline (days)", y = "Expected risk",
       colour = "CD59 expression") +
  theme_prism()
```

# 13. ssGSEA: Hallmark pathways, CD59-high vs -low

```{r ssgsea, eval=FALSE}
library(GSVA)
library(GSEABase)
library(pheatmap)

hallmark_sets <- getGmt(file.path(data_dir, "h.all.v2023.2.Hs.symbols.gmt"))

valid_samples <- rownames(survplotdata)[
  survplotdata$CD59quartile %in% c("Top 25", "Bottom 25")
]
expr_subset <- as.matrix(harm_data[, valid_samples, drop = FALSE])

# GSVA >= 1.50 uses the parameter-object API; older versions use arguments.
if (utils::packageVersion("GSVA") >= "1.50.0") {
  param <- ssgseaParam(exprData = expr_subset, geneSets = hallmark_sets)
  ssgsea_scores <- gsva(param)
} else {
  ssgsea_scores <- gsva(expr_subset, hallmark_sets,
                        method = "ssgsea", kcdf = "Gaussian")
}

# Order columns by CD59 group with a visual gap between the two blocks.
grp <- factor(survplotdata[colnames(ssgsea_scores), "CD59quartile"],
              levels = c("Top 25", "Bottom 25"))
ord <- order(grp)
ssgsea_sorted <- ssgsea_scores[, ord]

ann_col <- data.frame(CD59 = grp[ord])
rownames(ann_col) <- colnames(ssgsea_sorted)

pheatmap(
  ssgsea_sorted,
  scale = "row", cluster_cols = FALSE,
  annotation_col = ann_col,
  annotation_colors = list(CD59 = c("Top 25" = "red", "Bottom 25" = "blue")),
  gaps_col = sum(ann_col$CD59 == "Top 25"),
  color = colorRampPalette(c("blue3", "white", "red3"))(100),
  fontsize_row = 8, show_colnames = FALSE,
  main = "ssGSEA Hallmark scores by CD59 expression\n(Top 25% left, Bottom 25% right)"
)
```

# 14. Outcome-anchored ("reverse") analysis

Rather than splitting on CD59, define clear good- and poor-outcome groups and
ask which CD genes separate them.

```{r reverse-groups}
# dplyr verbs are namespace-qualified throughout sections 14-15: loading
# GSEABase/AnnotationDbi (sections 13, 15) masks select()/filter() with S4
# generics that do not operate on data frames.
reverse_df <- survplotdata %>%
  dplyr::mutate(
    outcome_group = dplyr::case_when(
      vitalStatus == 1 & overallSurvival <= 365  ~ "Poor",
      vitalStatus == 0 & overallSurvival >= 1095 ~ "Good",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(outcome_group))
reverse_df$outcome_group <- factor(reverse_df$outcome_group,
                                   levels = c("Good", "Poor"))
table(reverse_df$outcome_group)
```

```{r reverse-cd59}
ggplot(reverse_df, aes(x = outcome_group, y = CD59, fill = outcome_group)) +
  geom_boxplot(width = 0.25, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 1.4) +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  theme_classic(base_size = 14) +
  labs(x = "Outcome group", y = "CD59 (z-score)",
       title = "CD59: poor vs good outcome")
```

```{r reverse-screen}
reverse_hits <- compare_signatures(
  reverse_df, "outcome_group", cd_gene_cols,
  group1 = "Good", group2 = "Poor"
) %>%
  dplyr::filter(delta_mean > 0, FDR < 0.10)

head(reverse_hits)
```

# 15. AML cell-state signatures vs CD59 expression

Score each sample for published AML hierarchy / LSC signatures, then relate
those scores to CD59.

```{r celltype-scores, eval=FALSE}
library(GSEABase)

aml_gmt  <- getGmt(file.path(data_dir, "AMLCellType_Genesets.gmt"))
aml_sets <- geneIds(aml_gmt)

# Restrict each set to genes present in the expression matrix.
expr_genes <- rownames(zscore_counts)
aml_sets_matched <- lapply(aml_sets, function(gs) unique(gs[gs %in% expr_genes]))

# Signature score = mean z-score of member genes (NA-safe).
score_signature <- function(zmat, genes) {
  genes <- base::intersect(genes, rownames(zmat))
  if (length(genes) == 0) {
    return(setNames(rep(NA_real_, ncol(zmat)), colnames(zmat)))
  }
  if (length(genes) < 5) {
    warning("Fewer than 5 genes matched for a signature.")
  }
  colMeans(zmat[genes, , drop = FALSE], na.rm = TRUE)
}

aml_scores <- as.data.frame(
  lapply(aml_sets_matched, score_signature, zmat = zscore_counts),
  check.names = FALSE          # keep hyphenated signature names (e.g. "LSPC-Quiescent")
)
aml_scores <- tibble::tibble(sample_id = colnames(zscore_counts)) %>%
  dplyr::bind_cols(aml_scores)

sig_cols <- names(aml_sets_matched)
```

```{r celltype-compare, eval=FALSE}
aml_cd59_df <- survplotdata %>%
  tibble::rownames_to_column("sample_id") %>%
  dplyr::left_join(aml_scores, by = "sample_id") %>%
  dplyr::filter(CD59quartile %in% c("Bottom 25", "Top 25"))
aml_cd59_df$CD59quartile <- droplevels(aml_cd59_df$CD59quartile)

celltype_comparison <- compare_signatures(
  aml_cd59_df, "CD59quartile", sig_cols,
  group1 = "Bottom 25", group2 = "Top 25"
)
celltype_comparison

write.csv(celltype_comparison,
          "CD59_high_vs_low_AML_celltype_signatures.csv",
          row.names = FALSE)
```

```{r celltype-boxplots, eval=FALSE}
plot_df <- aml_cd59_df %>%
  dplyr::select(sample_id, CD59quartile, all_of(sig_cols)) %>%
  tidyr::pivot_longer(all_of(sig_cols), names_to = "signature", values_to = "score")
plot_df$signature <- factor(plot_df$signature,
                            levels = celltype_comparison$signature)

ggplot(plot_df, aes(x = CD59quartile, y = score, fill = CD59quartile)) +
  geom_boxplot(width = 0.35, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.12, size = 1.1, alpha = 0.55) +
  stat_compare_means(method = "wilcox.test", label = "p.format",
                     comparisons = list(c("Bottom 25", "Top 25")), size = 3) +
  facet_wrap(~ signature, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = c("Bottom 25" = "blue", "Top 25" = "red")) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 8)) +
  labs(x = "CD59 expression tertile", y = "Signature score",
       title = "AML cell-state signatures by CD59 expression")

ggsave("CD59_AML_celltype_signature_boxplots.pdf", width = 13, height = 15)
```

# 16. Session information

```{r session-info}
sessionInfo()
```
